---
title: "Intro to Sha Package for Sleep and Health Analysis"
author: "Yunrou Gong"
date: "`r Sys.Date()`"
output: rmarkdown::html_document

vignette: >
  %\VignetteIndexEntry{Sleep and Health Analysis with Sha}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Introduction
The *Sha* package is used for analysing sleep duration and other atrributes related to life health. The name **Sha**  is the abbreviation of Sleep and Health Analysis. We may have curious about weather sleep duration will be influenced by the weight, tempreture, activities, day of the week, and heart rate. Intuitively, sleep time and the other attributes are not that correlated. But, we could use this package to conduct detail analysis to answer intereting questions, which may be counter intuitive. Additionaly, I also wonder wheather having pet or not will affect the other features. Basicly, the activities will increase correspondingly by having a dog as you will have to do dog walking at least once a day. We can also conduct some interesting analysis to see how your pet will influence on your life, such as sleeping duration, body weight. 

# Description of Original Dataset 
In order to discover the attributes that may have some associate with our personal sleep time, I choose to collect of health data of myself via 'googlesheets', which I made it published so you can fetch this dataset via "googlesheets" package. I believe that the attributes set well represent my interets on the research question. The measurement of the attributes are also important, which will be detail mentioned below. 
```{r loaddata, echo=TRUE, message=FALSE, warning=FALSE}
library("googlesheets")
googlekey <- gs_url("https://docs.google.com/spreadsheets/d/1KdOw0FIxY2Kt8UgY16F3UMQSC7KJaBgtmewNZUGuLY8/edit#gid=0",lookup = FALSE,visibility = "public")
mydata <- gs_read(googlekey, skip = 3)
```
```{r data_view, echo=TRUE, results='asis'}
knitr::kable(head(mydata), format = "markdown", align = 'c')
```
```{r str_data, echo= FALSE, include= FALSE}
no_of_col <- dim(mydata)[2]
var_names <- colnames(mydata)
```

The collected  dataset contains `r no_of_col` variables related to one individual's sleep and health data and recoded each observation daily. 
The following is the variable explanation and how I measure them: 

1. `r var_names[1]`: in format of month/date/year started from Sep.11 2017
2. `r var_names[2]`: the name of the day during a week (Mon,Tue,Wed,Thu,Fri,Sat,Sun)
3. `r var_names[3]`: body weight in kg measured at erery morning when wake up using the same weighing machine
4. `r var_names[4]`: total sleep time (hour) per night measured by "Sleep Cycle"" app on iphone
5. `r var_names[5]`: daily average tempreture in Celsius from website https://www.wunderground.com/history/airport/KROC/2013/12/17/MonthlyHistory.html?&reqdb.zip=&reqdb.magic=&reqdb.wmo=
6. `r var_names[6]`: a measurement of activity in steps via iphone 'health' app
7. `r var_names[7]`: heart rate in beats per minute (bpm) measured every morning when wake up via "Sleep Cycle"" app on iphone
8. `r var_names[8]`: weather you have dog or not 

# Data Cleaning
### Generate a New Variable from column "DayofWeek"
Intuitively, weekday should have similar pattern on sleep duration compared to that of weekend. So we could generate a new variable name as "DayType" to classify each day as weekday or weekend. The more convinient way is to use library **tidyverse** as showing below.

```{r dataclean01, echo=TRUE, message=FALSE}
library(tidyverse)
mydata <- mydata %>% 
  mutate(DayType = ifelse(DayofWeek %in% c("Sat", "Sun"), "Weekend", "Weekday")) %>% 
  select(-DayofWeek) 
```

```{r schema, echo=TRUE, results='asis'}
#assign the variable name with units of measurement
sch <- c("Date", "Weight(kg)", "SleepDuration(hr)", "Tempreture(°C)", "Activity(steps)", 
         "HeartRate(bpm)","HavingDog", "DayType")
knitr::kable(head(mydata, 10), col.names = sch, align = "c")

```

Now, the following analysis are based on this dataset which is named as mydata. 

# Explotory Data Analysis

## Basic Statistics and useful information
Here, we are going to use "tidyverse" library to manupilate the "mydata" dataset and get the useful information.

```{r count, echo=TRUE }
n <- mydata %>%
  summarize(n = n())
firstday <- mydata$Date[1]
lastday <- mydata$Date[length(mydata$Date)]  
avg_st <- round(mean(mydata$SleepDuration),2)
percent_avg_st <- round(avg_st/24,4)*100
```

There are `r n` observations in the **mydata** dataset recoded from the date of `r firstday`  to `r lastday`. The average of my sleep time per day is approximately `r avg_st` hour, which account for `r percent_avg_st` % time a day. 

```{r count2, echo=TRUE }

count2 <- mydata %>% 
  group_by(DayType) %>%
  tally()

```
During this period, there are `r count2$n[1]` weekdays, `r count2$n[2]` weekends. 
We are going to find out the average sleep duration on weekday and on weekend and arrange them to see which group has longer sleep duration.
```{r sleep_by_day, echo=TRUE, results= 'asis'}
sleep_by_day <- mydata %>% 
  group_by(DayType) %>% 
  summarise(avg_sleep = mean (SleepDuration)) %>% 
  mutate(avg_sleep = round (avg_sleep, 2)) %>% 
  arrange(desc(avg_sleep))

knitr::kable(sleep_by_day, format = "markdown", align = 'l')

weekend_sd <- sleep_by_day$avg_sleep[1]
weekday_sd <- sleep_by_day$avg_sleep[0]
```
The average of sleep duration per day in weekend is ` r weekend_sd` hour, which is longer than the weekday sleep time on average, which is `r weekday_sd` hour based on our existing data.  It sounds very natually as I had no classes on weekend, so I may get up late. 

```{r sha, echo=TRUE, message=FALSE}
# Import the package *sha*
library(sha)
```

## Box plot 
We could also make box plot to visulise the differences of sleep duration between weekday and weekend by using  *plot_box* function. It will also show you how disperse the data is distributed , how skew the distribution may be, and also tell you the outliers in the dataset. 

```{r boxplot, echo=TRUE}
plot_box(mydata, x = "DayType", y= "SleepDuration")
```



To make the catergorical variable to factor variable for following analysis. 
```{r as.factor, echo=TRUE}
mydata$DayType <- as.factor(mydata$DayType)
mydata$HavingDog <- as.factor(mydata$HavingDog)

```


## Scatter plot
You can use *plot_scatter* function to make scatter plot on two continue variables by group to check if they have linear relationship and how does the relationship accross differnty groups. 
The below graph shows us the relationship between seelp duration and weight. It seems that there is some negative correlation  between sleep duration and body weight. 
```{r scatterplot, echo= TRUE}
plot_scatter(data = mydata,x = "Weight",y = "SleepDuration", color = "DayType")

```

Here is another example for looking at the relationship between activity steps and sleep duation.
You can use the *plot_scatter* function to make the plots on any two of the continue variables in the dataset, and grouped by a categorical vatiable. 

```{r scatterplot2, echo=TRUE}
plot_scatter(data = mydata,x = "Activity",y = "SleepDuration", color = "DayType")
```

From the above scatter plot, it seems that there is linear relationship between sleep duration and activity. And the slopes of weekday and weekend is very similar. 


### Checking statistically significantly mean difference across the groups
Now let's check weather there is statistically significantly difference on average sleep duration between weekday and weekend.
We are using oneway-anova test so it is made strong assumption that the dependent variable observation is normal distributed and do not have outliers.

**H0: The mean of sleep duration between weekday and weekend are equal**

**H1: The mean between the two groups are not equal**
```{r anova_test, echo=TRUE}
summary(aov(SleepDuration ~ DayType, data = mydata ))

```
From the above testing results, the p-value for this test is far less than the 0.05 significant level. So we reject the null hypothesis and conclude that there is significant difference in the average sleep duration between weekday and weekend under 95% confidence. 

We again did the oneway-anova test on the sleep duration between having a dog and not having a dog. 

**H0: The mean of sleep duration between having a dog or not are equal**

**H1: The mean between the two groups are not equal**

```{r, echo=TRUE}
summary(aov(SleepDuration ~ HavingDog, data = mydata))
```
Since the p-value is larger than the 0.05 significant level, we fail to reject the null hypotheis that the mean of sleep duration between having a dog or not having a dog are equal. So there is no significant differences of sleep duration on weather having a dog or not on average under given data and evidence , with 95 % confidence. 

I also wonder weather having a dog or not will lead to mean differences on my activity level. 

```{r anova_test2, echo=TRUE}
summary(aov(Activity~ HavingDog, data = mydata))
```

The p-value is far less than 0.05, which means there is siginifiant difference of activity steps on average weather having a dog or not. 

## Correlation Matrix Plot 

In the dataset, the weight , SleepDuration, Tempreture, Activity and Heart Rate are continues variables. So we can do a multiple correlation analysis on those continues variables to check the pair relationship. We can use the function *cor_plot* in "sha" package to calculate all the pair correlation among all the continues variable as *cor_plot* will automatically detect the continues variables and visulize the results in a plot. (see below)

```{r corrplot, echo=TRUE}
cor_plot(mydata)
```

From the above results, it is really interesting. The correlation between 

```{r, echo=TRUE,results='asis'}
cor.test(mydata$Weight,mydata$Activity)
```


Additionaly, we can use **tidyverse** library to calculate the correlation between two continues variables by set more condtions. For example, we are interested in the correlation between the sleep duration and the tempreture separately for each day type only for having a dog condition, and return a sorted list of those correlation coeffients rounded to two digits. 
```{r corrleration2, echo=TRUE, results='asis'}
ans <- mydata %>% 
  filter(HavingDog == "Yes") %>% 
  group_by(DayType) %>% 
  summarise(r = cor(SleepDuration, Tempreture)) %>% 
  mutate(r = round(r, 2)) %>% 
  arrange(desc(r))

knitr::kable(ans, format = "markdown", align = "l")  
```

So the results is quite interesting as there is positive correlation between sleep duration and tempreture on weekend and there is negative correlation between them on weekday when I have a dog. 



# Modelling 
### Simple Linear Regression Model
According to the above correlation results, we could conduct a simple linear regression models to quantify the association between heart rate and tempreture. 

```{r slm1, echo=TRUE}
slm_model <- slm_s(mydata, x = "Tempreture", y="HeartRate")
slm_model
modelplot(mydata, x = "Tempreture", y="HeartRate")

```

The equation of the above linear model can be constructed from the output:

**Predicted Heart Rate  = ** `r round(coef(slm_model)[1], 2)` **+ (**`r round(coef(slm_model)[2],2)`**) * Tempreture**

The interpretation for the slope coefficient: With a unit celsius increase in tempreture,the heart rate, on average, decrease by approximately 0.5 bpm.

The interpretation for the intercept coefficient: When the tempreture is 0,the heart rate, on average, is 87 bpm. 

There is  an approximation of 8.817 the residuas tend to deviate around the regression line.

However, only approximately 7% of the variability in the heart rate variable is explained by the tempreture variable.

The intercept is extremely significant, while the slop p value is larger than 5% significant level, 
So the overall model is not significant.

We can further check the model assumption by Q-Q plot to see if the data are normal distributed.
```{r normalitycheck, echo=TRUE}
qqnorm(slm_model$residuals)
qqline(slm_model$residuals)
```

It seemes that the data are not follow normal distribution, which violte the the **Normality** assumption. And the tail of the distribution has a lot of outliers, which may have huge impact on the shape of the distribution.


### Multiple Linear Regression Model
Create a scatter matrix plot on all the continues variables. 
```{r scattermatrix, echo=TRUE}
par(ps = 12, cex = 1, cex.main = 1)
plot(mydata %>% select(SleepDuration, Weight, Tempreture, Activity, HeartRate), pch=16, col="blue", main="Matrix Scatterplot of SleepDuration, Weight, Tempreture, Activity and HeartRate")
```

From the above graph, the variables seems that corrlated to each other , so it may arise multicollinearity problem. 


Creating a multiple linear regression model included all variables except the date, becasue the date is a distinct identifier, like index; and Weather is kind of overlapping the information that provided by "Tempreture". 

```{r ml01, echo=TRUE}
ml.saturated = lm(SleepDuration ~ Weight + Tempreture + Activity + HeartRate +  DayType + HavingDog, 
                  data= mydata)
mlr_sum <- summary(ml.saturated)
mlr_sum
knitr::kable(mlr_sum $coef, digits = c(4, 4, 3, 4), format = 'markdown')
```

The regression equation is :

**SleepDuration = 18 + (-0.22) * Weight + 0.03 * Tempreture - 0.0001 * Activity - 0.0029 * HeartRate + 1.55 * Weekend + 1.16 * HavingDog(Yes) **

Interpretion of the above equation under our problem context :

* When all attributes data are 0 and the day is weekend, the average expected duration of sleep is about 18 hour. 
* As the weight variable increases by one point, the sleep time decreases on average by about 0.22 hour (approximately 13 minute) while holding all other variables constant
* As the tempreture increases by one point, the sleep time increases on average by about 0.03 hour (approximately 2 minute) 
* As the activity increases by one step, the duration of sleep decreases on average by approximately 2 second
* As the heart rate increase by 1 bmp, the duration of sleep decreases on average by approximately 10 second
* The average difference in sleep duration between weekend and weekday is approximately 93 minute
* The average difference in sleep duration between having a dog and not having a dog is approximately 70 minutes

Although some of p-value of the attributes are less than 0.05, the overall regression is significant as the p-value is far less than 0.05. 

The RSE is about 1.455, which is an estimate of the average deviation of the observations around the regression line.

More importanly, the adjusted R square is 0.2913, meaning that approximately 29% of the variation in sleep duration is accounted for by the variables in our model, which indicates that the model is kind of poor in predicating the duration of sleep. Maybe, we should investigate the assumptions of the model. 




```{r assumption_check, echo= TRUE}
plot(ml.saturated, which = c(1,2))
```

According to the Q-Q plots, we will see that the data are follow the normal distributed. From the residual plot, we will also believe the data is normally distributed and it is independent. 



# Conclusion
Fitting linear models for predicting the duration of sleep seems do not have good performance based on the given data collected. And the data kind of not following normal distribution. But maybe because the size of the dataset is too small or we can use BoxCox to transform the data to see if the normality assumption is still violated.



```{r info, echo= T}
sessionInfo()
```


